---
title: "Low level R"
author: "Fabrice Rossi"
format: html
---

## R objects

R has variables that are used to access objects:

- functions
- data

### Functions

A function provides an action over objects or R itself. 

#### R related functions

```{r}
## this is the i_am function defined in the package here
## in general package::function
here::i_am("r-101-2024.Rproj")
## library loads the functions defined in a package (here)
library(here)
```

#### Functions that operate on objects

```{r}
class(1.0)
class(class)
```

If `f` is a function, `f(...)` applies the function to its parameters while 
`f` is the function as an object. 

## Data

Data objects can be separated into two main categories:

- vector: a finite set of values of a given type (like integers)
- list: a finite set of vectors (possibly of different types)

### Numerical vectors

```{r}
x <- 2
x
## x is a numeric vector
class(x)
## the type of x content is double
typeof(x)
```

```{r}
y <- 2L ## make clear to R that 2 is an integer (add L)
y
## y is a integer vector
class(y)
## the type of y content is integer
typeof(y)
```

```{r}
## square root calculation
z <- sqrt(y)
class(z)
typeof(z)
z
```

```{r}
## c is a construction function
x2 <- c(-2, 3, -1.5)
class(x2)
typeof(x2)
x2
```

Most of the functions operate in vectorized mode, i.e. their actions are applied
to each element of a vector independently from the other elements.

```{r}
## absolute value
y2 <- abs(x2)
y2
```

```{r}
sqrt(abs(x2))
```

```{r}
#| warning: false
sqrt(x2)
```

The special value `NaN` stands for *Not a Number* and is used when a numerical 
value cannot be associated to a mathematical expression. 

#### Numerical expressions

```{r}
x3 <- c(1, 2, -4)
y3 <- c(-1, 7, 42)
x3 + y3
```

```{r}
## vectorized product
x3 * y3
```

```{r}
z3 <- (x3 * y3) / (x3 + y3)
z3
```

```{r}
z3 + abs(z3)
z3 + x3
```

### Individual access

```{r}
## sequence from 0 to 5 (both included) with 0.5 increment (default 1)
x <- seq(0, 5, 0.5)
x
```

```{r}
length(x)
```

```{r}
## access to the third value in x
x[3]
```

```{r}
## out of range access gives NA
x[12]
```

The special value `NA` stands for `Not Available`. It is used for missing data
for instance.

```{r}
class(x[12])
typeof(x[12])
```

```{r}
## access to multiple values (sub-vector)
x[c(2, 4, 10)]
```

```{r}
## access to multiple values (sub-vector)
## negative values "remove" non wanted values
x[c(-2, -4, -10)]
```

```{r}
## a:b represents in a very efficient way the sequence from a to b (included) with
## 1 increment
x[4:8]
```

### Reusing values

```{r}
x1 <- 5
x2 <- seq(0, 4)
x1
x2
```

```{r}
x1 + x2 ## ??
```

```{r}
x2 + 1
```

```{r}
2 * x2 - 1
```

```{r}
1 / x2
```

```{r}
y <- seq(0, 5)
z <- c(1, -1)
y + z
```

```{r}
y <- seq(0, 4)
z <- c(1, -1)
y + z
```


